---
title: "Homework 2"
author: "Matthew T. Russell"
date: "10/9/2021"
output: github_document
---

```{r setup, load packages, include = FALSE}
library(tidyverse)
library(readxl)
library(readr)
```

# Problem 1

###### Reading and cleaning Mr. Trash Wheel sheet:
* specify the sheet in the Excel file and to omit non-data entries
(rows with notes/figures; columns containing notes) using arguments in `read_excel`
* use reasonable variable names
* omit rows that do not include dumpster-specific data
* round the number of sports balls to the nearest integer

```{r, read in Mr. Trash Wheel sheet and clean, message=FALSE}

trash_data <- 
  read_excel("./Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", 
  sheet = "Mr. Trash Wheel") %>% 
  janitor::clean_names() %>% 
  select(-x15, -x16, -x17) %>% 
  filter(!(str_detect(month, "Total"))) %>% 
  mutate(sports_balls = round(sports_balls, 0))

```

###### Reading and cleaning 2018 and 2019 precipitation sheets:
* omit rows without precipitation data
* add a variable for year
* combine precipitation datasets
* convert month to a character variable

```{r, read and clean precipitation sheets}

precip_2018_data <- 
  read_excel("./Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", 
  sheet = "2018 Precipitation", range = "A2:B14") %>% 
  janitor::clean_names() %>% 
  filter(!(is.na(total))) %>% 
  mutate(year = 2018)

precip_2019_data <- 
  read_excel("./Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", 
  sheet = "2019 Precipitation", range = "A2:B14") %>%
  janitor::clean_names() %>% 
  filter(!(is.na(total))) %>% 
  mutate(year = 2019)

precip_18_19_data <-
  bind_rows(precip_2018_data, precip_2019_data) %>% 
  mutate(month = month.name[month])
```

###### Summary of Mr. Trash Wheel and precipitation datasets:

There are `r nrow(trash_data)` observations and `r ncol(trash_data)` columns in the Mr. Trash Wheel dataset and `r nrow(precip_18_19_data)` observations and `r ncol(precip_18_19_data)` columns in the combined precipitation dataset. The Mr. Trash Wheel data tell us about the quantity of certain trash items such as plastic bottles, cigarette butts, glass bottles, grocery bags, chip bags, and sports balls. Meanwhile, the precipitation data tell us about the total rainfall (inches) per month for the years `r min(pull(precip_18_19_data))` and `r max(pull(precip_18_19_data))`. The median number of sports balls consumed by Mr. Trash Wheel is `r median(pull(trash_data, sports_balls))`. The total inches of rainfall in 2018 was `r sum(pull(precip_2018_data, total))`.



