---
title: "Homework 2"
author: "Matthew T. Russell"
date: "10/9/2021"
output: github_document
---

```{r setup, load packages, include = FALSE}
library(tidyverse)
library(readxl)
library(readr)
```

# Problem 1

###### Reading and cleaning Mr. Trash Wheel sheet:
* specify the sheet in the Excel file and to omit non-data entries
(rows with notes/figures; columns containing notes) using arguments in `read_excel`
* use reasonable variable names
* omit rows that do not include dumpster-specific data
* round the number of sports balls to the nearest integer

```{r, read in Mr. Trash Wheel sheet and clean, message=FALSE}

trash_data <- 
  read_excel("./Data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", 
  sheet = "Mr. Trash Wheel") %>% 
  janitor::clean_names() %>% 
  select(-x15, -x16, -x17) %>% 
  filter(!(str_detect(month, "Total"))) %>% 
  mutate(sports_balls = round(sports_balls, 0), 
         month = str_replace(month, "Decemeber", "December"))

```

###### Reading and cleaning 2018 and 2019 precipitation sheets:
* omit rows without precipitation data
* add a variable for year
* combine precipitation datasets
* convert month to a character variable

```{r, read and clean precipitation sheets}

precip_2018_data <- 
  read_excel("./Data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", 
  sheet = "2018 Precipitation", range = "A2:B14") %>% 
  janitor::clean_names() %>% 
  mutate(year = 2018)

precip_2019_data <- 
  read_excel("./Data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", 
  sheet = "2019 Precipitation", range = "A2:B14") %>%
  janitor::clean_names() %>% 
  mutate(year = 2019)

precip_18_19_data <-
  bind_rows(precip_2018_data, precip_2019_data) %>% 
  mutate(month = month.name[month])
```

###### Summary of Mr. Trash Wheel and precipitation datasets:

There are `r nrow(trash_data)` observations and `r ncol(trash_data)` columns in the Mr. Trash Wheel dataset. The median number of sports balls consumed by Mr. Trash Wheel in 2019 is `r trash_data %>% filter(year == 2019) %>% summarise(median(sports_balls))`. Key variables communicate type of trash Mr. Trash Wheel consumes, such as:

* `cigarette_butts`: number of cigarette butts
* `glass_bottles`: number of glass bottles
* `grocery_bags`: number of grocery bags
* `chip_bags`: number of bags of chips
* `sports_balls`: number of sports balls

Meanwhile, there are `r nrow(precip_18_19_data)` observations and `r ncol(precip_18_19_data)` columns in the combined precipitation dataset. The precipitation data tell us about the total rainfall (inches) per month for the years `r min(pull(precip_18_19_data))` and `r max(pull(precip_18_19_data))`.  The total inches of rainfall in 2018 was `r sum(pull(precip_2018_data, total))`.

# Problem 2

###### Reading and cleaning pols-months csv:
* clean data
* break up variable mon into integer variables year, month, and day
* replace month number with month name
* create a president variable taking values gop and dem
* remove prez_dem, prez_gop, and day variables

```{r, read/clean pols-month, message=FALSE}

pol_months_data <- read_csv(file = "./Data/pols-month.csv") %>% 
  separate(mon, into = c("year", "month", "day"), sep = "-", convert = T) %>% 
  mutate(month = month.name[month], 
         month = str_to_lower(month),
         president = prez_dem, 
         president = factor(president, levels = c(0,1), labels = c("gop", "dem"))) %>% 
  select(-c(prez_dem, prez_gop, day)) %>% 
  relocate(president, .after = month)

```

###### Reading and cleaning snp csv:
* arrange columns so year and month are first

```{r, read and clean snp, message=FALSE}

snp_data <- read_csv(file = "./Data/snp.csv") %>% 
  mutate(date = as.Date(date, "%m/%d/%y")) %>% 
  separate(date, into = c("year", "month", "day"), sep = "-", convert = T) %>% 
  mutate(month = month.name[month],
         month = str_to_lower(month),
         year = ifelse(year > 2021, year - 100, year)
        ) %>% 
  select(-day)

```

###### Tidying the unemployment csv:
* change data from 'wide' to 'long' format 

```{r, read and tidy unemployment, message=FALSE}
unemployment_data <- read_csv(file = "./Data/unemployment.csv") %>% 
  pivot_longer(
    Jan:Dec, 
    names_to = "month",
    values_to = "rate"
  ) %>% 
  mutate(
    month = match(month, month.abb), 
    month = month.name[month], 
    month = str_to_lower(month)
  ) %>% 
  janitor::clean_names()

```

###### Merge datasets:

```{r, merge datasets, message=FALSE}

pol_snp_unemp <- 
  left_join(pol_months_data, snp_data, by = c("year", "month")) %>% 
  left_join(unemployment_data, by = c("year", "month")) %>% 
  rename(unemp_rate = rate)
```

###### Summary of datasets:

The dataset from "pols-month" has provides information  about the number of republicans and democrats in congress as well the president's political party. The dataset from "snp" communicates the closing values of the Standard & Poor's stock market index. The dataset from "unemployment" gives the percent unemployed by month and year. After merging "snp" and "unemployment" data into "pols-month" data, we have `r nrow(pol_snp_unemp)` observations with `r ncol(pol_months_data)` columns about national political party representation, the health of the stock market, and the unemployment percentage from `r min(pull(pol_snp_unemp, year))` to `r max(pull(pol_snp_unemp, year))`. Key variables include:

* `president`: indicates whether president at that time was republican (gop) or democrat (dem)
* `close`: closing value of Standard & Poor's stock market index at that time
* `unemp_rate`: percent unemployed at that time

# Problem 3

###### Reading and tidying baby names csv:

```{r, read/tidy baby names, message=FALSE}

baby_names <- read_csv(file = "./Data/Popular_Baby_Names.csv") %>% 
  janitor::clean_names() %>% 
  mutate(
    across(gender:childs_first_name, str_to_lower), 
    ethnicity = ifelse(
      ethnicity %in% c("asian and pacific islander", "asian and paci"),
        "aapi", ethnicity
      ), 
    ethnicity = ifelse(
      ethnicity == "black non hisp", "black non hispanic", ethnicity
      ),
    ethnicity = ifelse(
      ethnicity == "white non hisp", "white non hispanic", ethnicity
      )
    ) %>% 
  distinct()

```

###### Creating table for popularity of 'Olivia' over time:

```{r, olivia popularity, message=FALSE}

olivia_pop <- baby_names %>% 
  filter(childs_first_name == "olivia") %>% 
  select(-c(gender, childs_first_name, count)) %>%
  arrange(year_of_birth) %>% 
  pivot_wider(
    names_from = "year_of_birth", 
    values_from = "rank"
  )

library(knitr)
kable(olivia_pop, caption = "Popularity of 'Olivia' for NYC babies by ethnicity from 2011 to 2016")

```

###### Creating table for most popular name for male children:

```{r, popular male names}

male_pop <- baby_names %>% 
  filter(rank == 1 & gender == "male") %>% 
  select(-c(gender, count, rank)) %>% 
  arrange(year_of_birth) %>% 
  mutate(childs_first_name = factor(childs_first_name)) %>% 
  pivot_wider(
    names_from = "year_of_birth", 
    values_from = "childs_first_name"
  ) %>% 
  mutate(across(as.character(2011:2016), as.character))

kable(male_pop, caption = "Most popular NYC male baby name by ethnicity from 2011 to 2016")

```

###### Creating a scatterplot of the number of male children with a name against the rank in popularity of that name among white non-Hispanic children born in 2016:

```{r, plot children name count vs name rank}

white_2016_names <- baby_names %>% 
  filter(ethnicity == "white non hispanic" & year_of_birth == 2016 & gender == "male")
  
ggplot(white_2016_names, aes(x = rank, y = count)) + geom_point() +
  ggtitle("Number of Male Children with a Name against Name's Rank in Popularity Among
          White Non-Hispanic NYC Children Born in 2016") +
  theme(plot.title = element_text(size = 12, hjust = 0.5))

```





